A Note on the Instability and Degeneracy of Deep Learning Models

### 2017-11-02 writing/paper-note/chapters/instab.Rmd
forgot to push last time

We consider the behavior, and the potential impropriety, sequences of discrete probability models built to incorporate observations of increasing sample size $N$.  Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.  The concept of instability was introduced in the field of statistical physics (i.e., point processes) by @ruelle1999statistical and then further extended by Schweinberger @schweinberger2011instability for a family of exponential models. At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation. As an example, such models can include near-degenerate distributions that assign all essential probability mass to only a subset of an overall sample space. The latter issue in connection to degeneracy has been recognized as concern in that dominant model outcomes  may not resemble observed data [cf. @handcock2003assessing]. As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions. Namely, volatilities in probability structure can potentially hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC). These reasons motivate our general study of instability for a broad class of probability models, described next.

@@

Section \@ref(examples) provides some examples of FOES models encountered in graph/network analysis and machine learning (i.e., deep learning models). These are used as references for later illustrations. Section \@ref(instability-results) then  establishes several formal results for  FOES models with regard to instability.  @schweinberger2011instability originally developed instability results framed for, or specific to, a certain class of discrete exponential models. For similar exponential models with random networks, @handcock2003assessing studied model degeneracy, where a probability model places near complete mass on  modes and may thereby narrow the feasible model outcomes. As findings here and from @schweinberger2011instability suggest, model instability and degeneracy may also be related by viewing degeneracy as an extreme, or limiting form, of instability. Our main results establish a broad characterization of model instability, appropriate across the whole FOES model class, that incorporates results of @schweinberger2011instability as a special case. We prescribe a general and simple condition  for identifying instability in a FOES model sequence, which  quantifies whether certain maximal probabilities in a FOES model  are too extreme relative to the sample size $N$. When fulfilled, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme peaks and troughs across nearly identical outcomes. As the measure of model instability increases, probabilities from an unstable FOES model additionally increase in volatility and provably slide into degeneracy. Section \@ref(implications) then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and   MCMC-based simulations. As one potential remedy, suggestions are given or constraining model parameterizations and avoid the most problematic regions of the parameter space. Proofs of the main results appear in Appendix \@ref(appendix-instab).

@@

## Discrete exponential family models

@@

with sufficient statistic $\boldsymbol g_N(\boldsymbol x)$ involving a count $\sum_{i=1}^N \mathbb{I}(x_i=j)$ for each outcome $j \in \{1,\ldots,k\}$, where $\mathbb{I}(\cdot)$ denotes the indicator function, and parameters $\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k$ defining log-probability ratios $\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]$. In addition to such standard models for discrete independent data,
exponential models of FOES type commonly arise with dependent spatial data [@besag1974spatial] and network/relational data [@wasserman1994social; @handcock2003assessing]. For a random graph  or network  with, say, $n$ nodes,  consider $N={n \choose 2}$ random edges where the $i$th edge is associated with a pair of nodes $s_i \equiv \{v_i,u_i\}$  and a binary variable $X_i\in\{0,1\}$ indicating   presence/absence of an edge among the node pair $s_i$, $i=1,\ldots,N$. Here the length $N$ of the edge variable sequence $\boldsymbol X = (X_1,\ldots,X_N)$ increases as a function of node number $n$ and corresponding exponential models often incorporate graph topographical features derived from $\boldsymbol X$. As an example, consider a   graph model of exponential/FOES form prescribed by

@@

## Restricted Boltzmann machines

@@

Here the baseline model \@ref(eq:RBM1) for hidden/visible variables is a linear exponential one in sufficient statistics $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ using $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ from   \@ref(eq:RBM1), but the form differs from the previous exponential models in \@ref(eq:expo) in that the lengths of parameters $\boldsymbol \theta_N$ and statistics  $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ increase to incorporate more visible variables. That is, in contrast to \@ref(eq:expo), the natural parameter function involved in the RBM model \@ref(eq:RBM1), as the identity mapping of the parameters $\boldsymbol \theta_N\in\mathbb{R}^{q(N)}$, naturally grows in  dimension  $q(N)\to \infty$ to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$. Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RMB model \@ref(eq:RBM1) to define a marginal model \@ref(eq:RBM2) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially   increase with $N$. Because $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM  specification \@ref(eq:RBM2) for visibles $\boldsymbol$ corresponds to a FOES model, while the joint distribution \@ref(eq:RBM1) for $(\boldsymbol X, \boldsymbol H)$ is also a FOES model. As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \@ref(eq:RBM1), is again a FOES model.

## Deep learning

@@

**Deep belief network (DBN).** A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers. The difference between the DBM and DBN models is that all but the last stacked layer in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs. Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N$, a DBN is also a FOES model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles. Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$, and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, in the definition of a FOES model.

# Main results on model instability {#instability-results}

## A criterion for instability {#criterion}

To define a measure of instability in FOES models, it is useful to consider the behavior of  data  models $P_{\theta_N}$, again supported on a set $\mathcal{X}^N$ of outcomes for $\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)$,  in connection to the sample size $N$.  A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
\begin{align}
(\#eq:elpr)
 \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
\end{align}
based on maximum and minimal model probabilities. In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for $N$ random variables where the log-ratio \@ref(eq:elpr) is overly large relative to the sample size $N$. That is, a sequence of FOES probability models $P_{\boldsymbol \theta_N}$ results in specifying the distribution of observations $\boldsymbol X=(X_1,\ldots,X_N)$ for each sample size $N \geq 1$ and instability will generally occur among these models whenever the corresponding LREP \@ref(eq:elpr) grows faster than $N$. This leads to the following definition.

@@

This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibility non-exponential family models (e.g., RBM and DBM models in Sections \@ref(restricted-boltzmann-machines) - \@ref(deep-learning) as well as a potentially increasing number $d(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{d(N)}$. While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf. Section \@ref(discrete-exponential-family-models) considered there. Section \@ref(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization. We next describe several potentially undesirable features associated with unstable models.

@@

## Characterizations and consequences of instability

As a basic characteristic, S-unstable FOES model sequences  have extremely sensitive probability structures. One aspect is that small changes in data configuration can   lead to very large changes in probability. Consider, for example, the quantity given by
$$
\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
$$
which represents the biggest log-probability ratio for a one component change in data outcomes in a FOES model with parameter $\boldsymbol \theta_N$.  We then have the following result prescribing the behavior of $\DN(\boldsymbol \theta_N)$ for unstable models.

@@

Theorem \@ref(thm:instab-elpr)(i) is a non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$. If so, Theorem \@ref(thm:instab-elpr)(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability.  Furthermore, and perhaps more seriously under Theorem \@ref(thm:instab-elpr)(ii),
S-unstable models can never have universally bounded changes in probability among single component variations in data configurations. While not all one component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.  As a consequence, unstable probability structure may exhibit extreme sensitivity through large peaks and troughs over the sample space.

Additionally, S-unstable FOES model sequences are also connected to degenerate models, where *degeneracy* involves assigning essentially all probability to modes within the sample space, which could potentially represent a  small subset among the totality of outcomes. For perspective, note that differing sizes of the scaled log-ratio $\REP(\boldsymbol \theta_N)/N$ from \@ref(eq:elpr) induce a spectrum of levels of instability/stability and Theorem \@ref(thm:instab-elpr) indicates increasing sensitivity of model probabilities as \@ref(eq:elpr) increases. Furthermore, as the instability measure grows and the log-ratio $\REP(\boldsymbol \theta_N)/N$ diverges, as in the definition \@ref(eq:Sun) of S-unstable models, then a FOES model sequence  will become degenerate. Theorem \@ref(thm:degenFOES) provides a formal statement of such degeneracy due to instability. For a given $0 < \epsilon < 1$, define a $\epsilon$-modal set of   outcomes as
\begin{equation}
(\#eq:mode)
\mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
\end{equation}

@@

Connected to degeneracy, unstable models may also exhibit additional kinds of  extreme and undesirable sensitivity in probabilities if model parameters $\boldsymbol \theta_N$ can further be "dialed" between positive and negative values. That is, some FOES models naturally involve parameter spaces covering a positive-negative spectrum of parameter possibilities, where the signs of parameters provide a standard device for increasing or decreasing probabilities of outcomes in the model formulation.  In fact, for many models, the switch of a parameter sign serves to produce reciprocal probabilities, as outlined in the following model assumption about parameter sign reversal (PSR).

@@

```{remark, name="need to briefly finish", echo=TRUE}
extensions of sensitivity... basically, model condition PSR and Corollary \ref{cor:sign} can be extended to cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ are not all changed in sign ($-\boldsymbol \theta_N$) but, more generally, are instead altered to another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$ involving a switch in sign only among some model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen. If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then an unstable model under $\boldsymbol \theta_N$ will then imply that many more unstable models exist  over a broad spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$.
```

@@

Model instability can depend intricately on how functions of parameters and data $\boldsymbol X_N=(X_1,\ldots,X_n)$ are combined in the formulation of the model probabilities, though some general causes may be identified. As one issue, a broad parameter space (or wide interpretation of this space) may admit some parameters as technically valid that have an undue and often undesirable impact on the model structure for a prescribed data size $N$. In this case, both the size and dimension of model parameters can be problematic and induce instability. In combination to this last point, further causes of instability may also be traced to the magnitude of statistics in the model. Potentially massive, and thereby unstable, statistics were the primary focus of instability studies of @schweinberger2011instability for certain discrete exponential models having parameters/statistics of fixed dimension. However, as shown in the following, bounded statistics may still lead to instability if the parameter dimension is high. We next provides some examples to illustrate instability in FOES models, which also suggest  some potential strategies for preventing unstable models.

## Equi-probability models

@@

## One-parameter exponential models {#one-param-exp}

@@

## Fixed-dimensional linear exponential models {#fixed-dim-exp}

@@

To see this, consider an even number $n>2$ of nodes and let $\boldsymbol x_0$ denote the data outcome in $\mathcal{X}^N \equiv \{0,1\}^N$ with all $N = {n \choose 2}$ edges being zero, let $\boldsymbol x_1$ denote the outcome with all edges being 1, and let $\boldsymbol x_2$ denote the edge configuration from dividing the nodes into two equal groups, with no edges within a group and all edges between the groups (so that no triangles exist in $\boldsymbol x_2$). Then, the $N$-scaled  log-ratio \@ref(eq:elpr) for the exponential graph model \@ref(eq:mod2) can, by definition, be bounded below by
\begin{align*}
\frac{1}{N}\REP_N(\boldsymbol \theta_N) &\geq \max_{i=1,2}\frac{1}{N}
\left| \log\left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_0)}\right] \right| \\
&= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
\end{align*}
a similar expression also holds for an odd node number $n>2$. Consequently, for all fixed parameters excluding $\theta_2=\theta_3=0$, $\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty$ then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \@ref(prp:prop1) for this model when $|\theta_2|+|\theta_3|\neq 0$. That is, instability holds even under $\theta_2 = - \theta_3/3$ case potentially allowed by Schweinberger's results.

## Latent variable models of increasing parameter dimension

>>>>

We consider the behavior, and the potential impropriety, of sequences of discrete probability models built to incorporate observations of increasing sample size $N$. Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.  The concept of instability was introduced in the field of statistical physics (i.e., point processes) by @ruelle1999statistical and then further extended by Schweinberger @schweinberger2011instability for a family of exponential models. At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation. As an example, such models can include near-degenerate distributions that assign all essential probability mass to only a subset of an overall sample space. The latter issue in connection to degeneracy has been recognized as a concern in that dominant model outcomes  may not resemble observed data [cf. @handcock2003assessing]. As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions. Namely, volatilities in probability structure can potentially hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC). These reasons motivate our general study of instability for a broad class of probability models, described next.

@@

Section \@ref(examples) provides some examples of FOES models encountered in graph/network analysis and machine learning (i.e., deep learning models). These are used as references for later illustrations. Section \@ref(instability-results) then establishes several formal results for  FOES models with regard to instability.  @schweinberger2011instability originally developed instability results specific to a certain class of discrete exponential models. For similar exponential models with random networks, @handcock2003assessing studied model degeneracy, where a probability model places near complete mass on modes and may thereby narrow the feasible model outcomes. As findings here and from @schweinberger2011instability suggest, model instability and degeneracy may also be related by viewing degeneracy as an extreme, or limiting form, of instability. Our main results establish a broad characterization of model instability, appropriate across the whole FOES model class, that incorporates results of @schweinberger2011instability as a special case. We prescribe a general and simple condition for identifying instability in a FOES model sequence, which  quantifies whether certain maximal probabilities in a FOES model  are too extreme relative to the sample size $N$. When fulfilled, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme peaks and troughs across nearly identical outcomes. As the measure of model instability increases, probabilities from an unstable FOES model additionally increase in volatility and provably slide into degeneracy. Section \@ref(implications) then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and   MCMC-based simulations. As one potential remedy, suggestions are given or constraining model parameterizations to avoid the most problematic regions of the parameter space. Proofs of the main results appear in Appendix \@ref(appendix-instab).

@@

## Discrete Exponential Family Models

@@

with sufficient statistic $\boldsymbol g_N(\boldsymbol x)$ involving a count $\sum_{i=1}^N \mathbb{I}(x_i=j)$ for each outcome $j \in \{1,\ldots,k\}$, where $\mathbb{I}(\cdot)$ denotes the indicator function, and parameters $\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k$ defining log-probability ratios $\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]$. In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data [@besag1974spatial] and network/relational data [@wasserman1994social; @handcock2003assessing]. For a random graph  or network  with, say, $n$ nodes,  consider $N={n \choose 2}$ random edges where the $i$th edge is associated with a pair of nodes $s_i \equiv \{v_i,u_i\}$  and a binary variable $X_i\in\{0,1\}$ indicating   presence/absence of an edge among the node pair $s_i$, $i=1,\ldots,N$. Here the length $N$ of the edge variable sequence $\boldsymbol X = (X_1,\ldots,X_N)$ increases as a function of node number $n$ and corresponding exponential models often incorporate graph topographical features derived from $\boldsymbol X$. As an example, consider a graph model of exponential/FOES form prescribed by

@@

## Restricted Boltzmann Machines

@@

Here the baseline model \@ref(eq:RBM1) for hidden/visible variables is a linear exponential one in sufficient statistics $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ using $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ from   \@ref(eq:RBM1), but the form differs from the previous exponential models in \@ref(eq:expo) in that the lengths of parameters $\boldsymbol \theta_N$ and statistics  $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ increase to incorporate more visible variables. That is, in contrast to \@ref(eq:expo), the natural parameter function involved in the RBM model \@ref(eq:RBM1), as the identity mapping of the parameters $\boldsymbol \theta_N\in\mathbb{R}^{q(N)}$, naturally grows in  dimension  $q(N)\to \infty$ to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$. Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RBM model \@ref(eq:RBM1) to define a marginal model \@ref(eq:RBM2) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially increase with $N$. Because $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM  specification \@ref(eq:RBM2) for visibles $\boldsymbol$ corresponds to a FOES model, while the joint distribution \@ref(eq:RBM1) for $(\boldsymbol X, \boldsymbol H)$ is also a FOES model. As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \@ref(eq:RBM1), is again a FOES model.

## Deep Learning

<!-- http://search.library.duke.edu/search?id=DUKE002216318 -->

@@

**Deep belief network (DBN).** A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers. The difference between the DBM and DBN models is that all but the last stacked layer in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs. A Bayesian network is a probabilistic graphical model that defines conditional dependence to be directed, rather than undirected (as with the RBM).  Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N, \mid \mathcal{X} \mid < \infty$, a DBN is also a FOES model with $q(N)$ the length of parameter vector is dependent on the dimension of the visibles because $P_{\boldsymbol \theta_N}(\boldsymbol x)>0$ for all $\boldsymbol x \in\mathcal{X}^N$. Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$, and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, satisfying the definition of a FOES model.

# Main Results on Model Instability {#instability-results}

We now present a formal definition for instability of FOES models as well as a simple condition for identifying instability in a FOES model sequence.

## A Criterion for Instability {#criterion}

To define a measure of instability in FOES models, it is useful to consider the behavior of data  models $P_{\theta_N}$, again supported on a set $\mathcal{X}^N$ of outcomes for $\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)$,  in connection to the sample size $N$.  A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
\begin{align}
(\#eq:elpr)
 \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
\end{align}
based on maximum and minimal model probabilities. In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for $N$ random variables where the LREP \@ref(eq:elpr) is overly large relative to the sample size $N$. That is, a sequence of FOES probability models $P_{\boldsymbol \theta_N}$ results in specifying the distribution of observations $\boldsymbol X=(X_1,\ldots,X_N)$ for each sample size $N \geq 1$ and instability will generally occur among these models whenever the corresponding LREP \@ref(eq:elpr) grows faster than $N$. This leads to the following definition.

@@

This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) as well as a potentially increasing number $q(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{q(N)}$. While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf. Section \@ref(discrete-exponential-family-models) considered there. Section \@ref(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization. We next describe several potentially undesirable features associated with S-unstable FOES models.

@@

## Characterizations and Consequences of Instability

As a basic characteristic, S-unstable FOES model sequences  have extremely sensitive probability structures. One aspect is that small changes in data configuration can   lead to very large changes in probability. Consider, for example, the quantity given by
$$
\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
$$
which represents the biggest log-probability ratio for a one component change in data outcomes in a FOES model with parameter $\boldsymbol \theta_N$.  We then have the following result prescribing the behavior of $\DN(\boldsymbol \theta_N)$ for S-unstable FOES models.

@@

Theorem \@ref(thm:instab-elpr)(i) is a non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$. If so, Theorem \@ref(thm:instab-elpr)(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability. Furthermore, and perhaps more seriously under Theorem \@ref(thm:instab-elpr)(ii),
S-unstable models can never have universally bounded changes in probability among single component variations in data configurations. While not all one component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property. As a consequence, unstable probability structures may exhibit extreme sensitivity through large peaks and troughs over the sample space.

Additionally, S-unstable FOES model sequences are also connected to degenerate models, where *degeneracy* involves assigning essentially all probability to modes within the sample space, which could potentially represent a  small subset among the totality of outcomes. For perspective, note that differing sizes of the scaled log-ratio $\REP(\boldsymbol \theta_N)/N$ from \@ref(eq:elpr) induce a spectrum of levels of instability/stability and Theorem \@ref(thm:instab-elpr) indicates increasing sensitivity of model probabilities as \@ref(eq:elpr) increases. Furthermore, as the instability measure grows and the log-ratio $\REP(\boldsymbol \theta_N)/N$ diverges, as in the definition \@ref(eq:Sun) of S-unstable models, then a FOES model sequence  will become degenerate. Theorem \@ref(thm:degenFOES) provides a formal statement of such degeneracy due to S-instability. For a given $0 < \epsilon < 1$, define a $\epsilon$-modal set of outcomes as
\begin{equation}
(\#eq:mode)
\mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
\end{equation}

@@

Connected to degeneracy, S-unstable FOES models may also exhibit additional kinds of  extreme and undesirable sensitivity in probabilities if model parameters $\boldsymbol \theta_N$ can further be "dialed" between positive and negative values. That is, some FOES models naturally involve parameter spaces covering a positive-negative spectrum of parameter possibilities, where the signs of parameters provide a standard device for increasing or decreasing probabilities of outcomes in the model formulation.  In fact, for many models, the switch of a parameter sign serves to produce reciprocal probabilities, as outlined in the following model assumption about parameter sign reversal (PSR).

@@

```{remark, echo=TRUE}
Model condition PSR and Corollary \ref{cor:sign} can be extended to cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ are not all changed in sign ($-\boldsymbol \theta_N$) but, more generally, are instead altered to another parameter configuration $\boldsymbol \theta_{N}^A = (\boldsymbol \theta_{N,1}^A, \boldsymbol \theta_{N,2}^A)$ involving a switch in sign only among some model parameters $\boldsymbol \theta_{N,2}^A = -\boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen. If a sign change occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then an unstable model under $\boldsymbol \theta_N$ will then imply that many more unstable models exist over a broad spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$.
```

@@

Model instability can depend intricately on how functions of parameters and data $\boldsymbol X_N=(X_1,\ldots,X_n)$ are combined in the formulation of the model probabilities, though some general causes may be identified. As one issue, a broad parameter space (or wide interpretation of this space) may admit some parameters as technically valid that have an undue and often undesirable impact on the model structure for a prescribed data size $N$. In this case, both the size and dimension of model parameters can be problematic and induce instability. In combination to this last point, further causes of instability may also be traced to the magnitude of statistics in the model. Potentially massive, and thereby unstable, statistics were the primary focus of instability studies of @schweinberger2011instability for certain discrete exponential models having parameters/statistics of fixed dimension. However, as shown in the following, bounded statistics may still lead to instability if the parameter dimension is high. We next provides some examples to illustrate S-instability in FOES models, which also suggest some potential strategies for preventing unstable models.

## Equi-probability Models

@@

## One-parameter Exponential Models {#one-param-exp}

@@

## Fixed-dimensional Linear Exponential Models {#fixed-dim-exp}

@@

To see this, consider an even number $n>2$ of nodes and let $\boldsymbol x_0$ denote the data outcome in $\mathcal{X}^N \equiv \{0,1\}^N$ with all $N = {n \choose 2}$ edges being zero, let $\boldsymbol x_1$ denote the outcome with all edges being 1, and let $\boldsymbol x_2$ denote the edge configuration from dividing the nodes into two equal groups, with no edges within a group and all edges between the groups (so that no triangles exist in $\boldsymbol x_2$). Then, the $N$-scaled  log-ratio \@ref(eq:elpr) for the exponential graph model \@ref(eq:mod2) can, by definition, be bounded below by
\begin{align*}
\frac{1}{N}\REP_N(\boldsymbol \theta_N) &\geq \max_{i=1,2}\frac{1}{N}
\left| \log\left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_0)}\right] \right| \\
&= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
\end{align*}
a similar expression also holds for an odd node number $n>2$. Consequently, for all fixed parameters excluding $\theta_2=\theta_3=0$, $\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty$ then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \@ref(prp:prop1) for this model when $|\theta_2|+|\theta_3|\neq 0$. That is, instability holds even under $\theta_2 = - \theta_3/3$ case potentially allowed by Schweinberger's [-@schweinberger2011instability] results.

## Latent Variable Models of Increasing Parameter Dimension

We next consider instability of discrete data models based on exponential formulations involving hidden, or latent, variables, such as those probabilistic graphical models described in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning). We will focus on restricted Bolzmann machine (RBM) models (Section \@ref(restricted-boltzmann-machines), having one layer of latent variables for simplicity, though the same instability concepts may be extended to other deep learning models (Section \@ref(deep-learning). For $N$ visible variables $\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)$ as data, each observation $X_i\in\{\pm 1\}$ being binary, the RBM-based model \@ref(eq:RBM2) for  $\boldsymbol X$ is again of FOES-type, though not an exponential model. However, the distribution of visible variables is induced by an underlying joint exponential model \@ref(eq:RBM1) for both visible and latent variables $(\boldsymbol X, \boldsymbol H)$, where $\boldsymbol H = (H_1,\ldots,H_{N_{\mathcal{H}}})$ denotes a vector of $N_{\mathcal{H}}$ hidden variables (similarly binary). The joint model is of linear exponential form involving $q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}$ sufficient statistics given by $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$ and parameters $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}  ) \in\mathbb{R}^{q(N)}$ corresponding to the $N$ visible variables $\boldsymbol X$ (i.e., $\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N$), the $N_{\mathcal{H}}$ hidden variables $\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}$), and the $N *N_{\mathcal{H}}$ cross-product variables $\boldsymbol X^T\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}$). However, unlike some previous exponential models considered in Sections \@ref(one-param-exp)-\@ref(fixed-dim-exp) (cf. Proposition \@ref(prp:prop1), note that the RBM formulation always associates parameters with *bounded* statistics (i.e., the components of $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$) so that  model instability cannot arise here due to the magnitude of sufficient statistics exceeding the sample size $N$. Instead, RBM instability  may be linked solely to parameter configuration and the fact that the number $q(N) \geq N$ of parameters necessarily increases with the number $N$ of observations $\boldsymbol X$, in contrast to previous exponential cases of fixed parameter dimension.
